---
title: "wilkinson"
author: "JMA"
date: "September 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## An exmple of numerical instability with Wilkinson's polynomial


Evaluate the Wilkinson polynomial by term by term construction -
$w_n(x) = \prod_{k=1\ldots n}(x-k)$
This avoids integer rounding errors -  a smoother version

```{r wilkinson.products}

wilkinson.products <- function(x, nn=20) {
    w <- function(x, n=nn) {
        v <- rep(x, n) - seq(n)  # The terms in the polynomial product (x-1), (x-2), ...
        Reduce( '*', v)
    }
sapply(x, w)
}
```

Expanding the polynomial, one finds
Note this suffers from a few percent noise due to integer overflow
```{r wilkinson.20}
wilkinson.20 <- function(x) {
    y <- x^20 - 210L * x^19 + 20615L * x^18 - 1256850L * x^17 + 53327946L * x^16 -
    1672280820L * x^15 +40171771630L * x^14 - 756111184500L * x^13 +
    11310276995381L * x^12 - 135585182899530L * x^11 +
    1307535010540395L * x^10 - 10142299865511450L * x^9 +
    63030812099294896L * x^8 - 311333643161390640L * x^7 +
    1206647803780373360L * x^6 - 3599979517947607200L * x^5 +
    8037811822645051776L * x^4 - 12870931245150988800L * x^3 +
    13803759753640704000L * x^2 - 8752948036761600000L * x +
    2432902008176640000L
    y
}
```

Google's _Protocol Buffers_ convert language-specific data structures to a universal binary transfer format. It has interfaces in several languages, but not `R`. Fortunately `RProtoBuf` wraps _Protocol Buffers_, and goes a step further, to generate protocol payload conversion descriptions (as contained in `.proto` file) dynamically, hiding the machinery needed with other languages. 

```{r}
if (!("RProtoBuf"  %in% .packages(all=TRUE)) ) {
    install.packages("RProtoBuf")
}
library(RProtoBuf)
```

On the Mac, for example, you may instead get an error during install:

>   checking google/protobuf/stubs/common.h usability... no
>   checking google/protobuf/stubs/common.h presence... no
>   checking for google/protobuf/stubs/common.h... no
>   configure: WARNING: Protobuf headers not found with default CXXFLAGS and CPPFLAGS, manually trying /usr/local/include

The `RProtobuf` install requires your machine already has Google's _Protocol Buffers_ installed. To find out if it's there already, see if the compiler `protoc` runs at the command line:
```{bash}
protoc --version
```
If instead the command isn't recognized, you need to install [_Protocol Buffers_](https://developers.google.com/protocol-buffers/) from [its github repository](https://github.com/google/protobuf.git). Be forewarned that you may discover that this takes more than just a few steps since it requires an extensive toolchain. _Do you know the story [If you give a Mouse a Cookie?]()._

Note: On the Mac you will need to have installed the _command line tools_ that come with Apple's _XCode_ development environment. You will also need the following:  `pkg-config`, which autoconf
autoconf
libtool
automake
glibtool (a replacement for Mac's native libtool)

/usr/bin/ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)"
brew install bazel
  
and [`bazel`]()
which can be installed with Mac package manager `brew` (available at ??)
Look at the `README.md` files in the root and `src` directories in the `protoBuf` repository for details. 



